---
title: "Cluster and annotation assessment via UMAP plots"
subtitle: "scRNA-seq of metabolite-treated T-cells"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
params:
  config: config.yaml
  date: !r Sys.Date()
---

# Notes

```{r chunk_settings, echo=FALSE}

knitr::opts_chunk$set(fig.dim = c(10, 10))

```

```{r date, echo=FALSE}

#params <- list(config = "config.yaml", date = Sys.Date())

today <- params$date
sprintf('Report run on %s.', today)

```

```{r load_libraries, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

libload <- c("Seurat", "future", "stringr", "dplyr", "ggplot2", "knitr", "yaml", "scales", "viridis", "patchwork", "ggrastr")

for (i in libload) {
    library(i, character.only = T)
}

# Custom functions

source("functions.R")

```

```{r setup, echo=FALSE, warning=FALSE, message=FALSE }

set.seed(7)

root_directories <- get_root_directories()
analysis_dir <- root_directories[["analysis"]]

config <- yaml.load_file(params$config)

# parallelization in Seurat with future
# see https://satijalab.org/seurat/articles/future_vignette.html
plan("multisession", workers = 6)
options(future.globals.maxSize= 2000*1024^2) #otherwise error that future max size is exceeded

```

```{r analysis_params, eval=T ,echo=F}

annotation_key <- "annotation_T-cell-subtypes"

clusterCol <- config$scParam$clusters
annotCol <- config[[annotation_key]]$annotation_col
condCol <- "group"
libCol <- "lib"

```

```{r load_data, eval=T ,echo=F}

SObj <- readRDS(file.path(analysis_dir, config[["Seurat_filtered"]]))

```

```{r split_seurat_object, eval=T ,echo=F}

SObj_split <- SplitObject(SObj, split.by = condCol)

```

# UMAP plots of annotation and clusters

## UMAP by annotation

```{r umap_annots, eval=T, echo=F, fig.dim = c(8, 7)}

umap_annot <- DimPlot(SObj, group.by = annotCol, reduction = config$scParam$dimred)

umap_annot

```

## UMAP by annotation and condition

```{r umap_annots_cond, eval=T, echo=F, fig.dim = c(12, 12)}

umap_annot_cond <- DimPlot(SObj, group.by = annotCol, split.by = condCol, reduction = config$scParam$dimred, ncol = 2)

umap_annot_cond

```

## UMAP by annotation and library

```{r umap_annots_lib, eval=T, echo=F, fig.dim = c(12, 12)}

umap_annot_lib <- DimPlot(SObj, group.by = annotCol, split.by = libCol, reduction = config$scParam$dimred, ncol = 2)

umap_annot_lib

```

## UMAP by annotation and condition, separate libraries {.tabset}

```{r umap_annots_cond_lib, eval=T, echo=F, results='asis', fig.dim = c(12, 12)}

umap_annot_cond_lib <- list()

for (cond in names(SObj_split)) {
    umap_annot_cond_lib[[cond]] <- DimPlot(
        SObj_split[[cond]],
        group.by = annotCol, split.by = libCol,
        reduction = config$scParam$dimred, ncol = 2
    )

    cat("### ", cond, "\n")
    print(umap_annot_cond_lib[[cond]])
    cat("\n\n")
}

```

## UMAP by clusters

```{r umap_clusters, eval=T, echo=F, fig.dim = c(8, 7)}

umap_clusters <- DimPlot(SObj, group.by = clusterCol, reduction = config$scParam$dimred)

umap_clusters

```

## UMAP by clusters and condition

```{r umap_clusters_cond, eval=T, echo=F, fig.dim = c(12, 12)}

umap_clusters_cond <- DimPlot(SObj, group.by = clusterCol, split.by = condCol, reduction = config$scParam$dimred, ncol = 2)

umap_clusters_cond

```

## UMAP by clusters and library

```{r umap_clusters_lib, eval=T, echo=F, fig.dim = c(12, 12)}

umap_clusters_lib <- DimPlot(SObj, group.by = clusterCol, split.by = libCol, reduction = config$scParam$dimred, ncol = 2)

umap_clusters_lib

```

## UMAP by clustering and condition, separate libraries {.tabset}

```{r umap_clusters_cond_lib, eval=T, echo=F, results='asis', fig.dim = c(12, 12)}

umap_clusters_cond_lib <- list()

for (cond in names(SObj_split)) {
    umap_clusters_cond_lib[[cond]] <- DimPlot(
        SObj_split[[cond]],
        group.by = clusterCol, split.by = libCol,
        reduction = config$scParam$dimred, ncol = 2
    )

    cat("### ", cond, "\n")
    print(umap_clusters_cond_lib[[cond]])
    cat("\n\n")
}

```

# QC UMAP plots

## Counts, features percent mitochondrial/ribo

```{r QC_counts, eval=T, echo=F, fig.dim = c(12, 9)}

qc_features <- c(
    "nCount_RNA",
    "nFeature_RNA",
    "percent.mt",
    "percent.ribo"
)

umap_qc <- list()

for (feature in qc_features) {
    umap_qc[[feature]] <- FeaturePlot(
        SObj, features = feature,
        reduction = config$scParam$dimred
    ) + scale_color_viridis()
}

umap_qc_combined <- wrap_plots(umap_qc)

umap_qc_combined

```

## Cell cycle

```{r QC_cell_cycle_1, eval=T, echo=F, fig.dim = c(12, 9)}

cc_features <- c(
    "S.Score",
    "G2M.Score",
    "S_UCellModule",
    "G2M_UCellModule"
)

umap_cc <- list()

for (feature in cc_features) {
    umap_cc[[feature]] <- FeaturePlot(
        SObj, features = feature,
        reduction = config$scParam$dimred
    ) + scale_color_viridis()
}

umap_cc_combined <- wrap_plots(umap_cc)

umap_cc_combined

```

```{r QC_cell_cycle_2, eval=T, echo=F, fig.dim = c(12, 5)}

umap_phase <- DimPlot(
    SObj, group.by = "Phase", reduction = config$scParam$dimred
)

umap_phase_smooth <- DimPlot(
    SObj, group.by = "Phase_smooth", reduction = config$scParam$dimred
)

umap_phase_with_smooth <- wrap_plots(umap_phase, umap_phase_smooth, ncol = 2)

umap_phase_with_smooth

```

## Computational doublets

```{r QC_comp_doublets, eval=T, echo=F, fig.dim = c(12, 5)}

umap_doublet_class <- DimPlot(
    SObj, group.by = "scDblFinder.class", reduction = config$scParam$dimred
)

umap_doublet_score <- FeaturePlot(
    SObj, features = "scDblFinder.score", reduction = config$scParam$dimred
)

umap_doublets <- wrap_plots(umap_doublet_class, umap_doublet_score, ncol = 2)

umap_doublets

```

## Signal-to-noise

Signal-to-noise ratio based on principal components

```{r QC_signal_to_noise, eval=T, echo=F, fig.dim = c(12, 5)}

pca_data <- t(Embeddings(SObj, reduction="pca"))

SObj[[]]$signal_to_noise_PCA_10 <- apply(pca_data, 2, function(x) sum(abs(x[1:10]))/sum(abs(x[11:50])))
SObj[[]]$signal_to_noise_PCA_20 <- apply(pca_data, 2, function(x) sum(abs(x[1:20]))/sum(abs(x[21:50])))

umap_signal_to_noise_10 <- FeaturePlot(
    SObj, features = "signal_to_noise_PCA_10",
    reduction = config$scParam$dimred
) + scale_color_viridis()

umap_signal_to_noise_20 <- FeaturePlot(
    SObj, features = "signal_to_noise_PCA_20",
    reduction = config$scParam$dimred
) + scale_color_viridis()

umap_signal_to_noise <- wrap_plots(umap_signal_to_noise_10, umap_signal_to_noise_20, ncol = 2)

umap_signal_to_noise

```

```{r save_plots, eval=F, echo=F}

plotPathBase <- file.path(analysis_dir, config$outputpath$fig)
plotFileType <- "pdf"

ggsave(
    file.path(plotPathBase, "umap_clusters", paste0("umap_clusters.", plotFileType)),
    rasterize(umap_clusters, layer = "point", dpi = 300) + labs(x = "UMAP 1", y = "UMAP 2"),
    width = 9, height = 8
)

ggsave(
    file.path(plotPathBase, "umap_clusters", paste0("umap_clusters_cond.", plotFileType)),
    rasterize(umap_clusters_cond, layer = "point", dpi = 300) + labs(x = "UMAP 1", y = "UMAP 2"), width = 12, height = 12
)

ggsave(
    file.path(analysis_dir, config$outputpath$fig, "interactive", "umap_annot_lib.png"),
    umap_annot_lib, width = 12, height = 7
)

ggsave(
    file.path(analysis_dir, config$outputpath$fig, "interactive", "umap_clusters_cond.png"),
    umap_clusters_cond, width = 12, height = 10
)

ggsave(
    file.path(analysis_dir, config$outputpath$fig, "interactive", "umap_clusters_lib.png"),
    umap_clusters_lib, width = 12, height = 7
)

ggsave(
    file.path(analysis_dir, config$outputpath$fig, "interactive", "umap_annot.png"),
    umap_annot, width = 10, height = 7
)

ggsave(
    file.path(analysis_dir, config$outputpath$fig, "interactive", "umap_clusters.png"),
    umap_clusters, width = 8, height = 7
)

ggsave(
    file.path(analysis_dir, config$outputpath$fig, "interactive", "umap_phase.png"),
    umap_phase_with_smooth, width = 12, height = 5
)

```

# SessionInfo

```{r sessionInfo}

sessionInfo()

```

