---
title: "Dimensionality Reduction"
author: "Malte Simon and Paul Heinrich, adapted from Inma Hernandez & Nicholas Strieder"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
params:
  config: config.yaml
  date: !r Sys.Date()
---


```{r date, echo=FALSE}

today <- params$date
sprintf('Report run on %s.', today)

```

```{r load libraries, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# libraries

libload<-c("rprojroot", "jsonlite", "yaml", "Seurat", "SeuratWrappers", "ggpubr", "tidyverse", "plyr", 
        "cowplot", "patchwork", "ggExtra","harmony", "UCell", "future", 'jj', 'viridis', 'ggpointdensity')
invisible(sapply(libload, library, character.only = T))

# Custom functions

source("functions.R")

```

```{r setup, echo=FALSE, warning=FALSE}

set.seed(7)

root_directories <- get_root_directories()
analysis_dir <- root_directories[["analysis"]]

config <- yaml.load_file(params$config)
config_schema <- read_json(file.path("schema", "config.schema.json"))

# parallelization in Seurat with future
# see https://satijalab.org/seurat/articles/future_vignette.html
plan("multisession", workers = 6)
options(future.globals.maxSize= 2000*1024^2) #otherwise error that future max size is exceeded

# bibliography
try(knitr::write_bib(c(.packages()), file.path(analysis_dir, config$outputpath$report, "R.bib")))

```

```{r parameters, echo=FALSE}

integration_method_wrappers <- list(
  "harmony" = HarmonyIntegration,
  "fastMNN" = FastMNNIntegration,
  "rpca" = RPCAIntegration,
  "cca" = CCAIntegration,
  "jpca" = JointPCAIntegration
)

```

```{r process_config, echo=FALSE}

check_integration_config(config, config_schema, analysis_dir)
check_clustering_config(config)

integrate <- config$integration$integrate
integration_methods <- c()
if (integrate) {integration_methods <- config$integration$methods}

cluster_resolution <- as.numeric(str_extract(config$scParam$clusters, "([0-9]+\\.[0-9]+$)|([0-9]+)$"))

```

```{r custom functions, include=F}

create_dimset_table <- function(dim_from = c(1,2), dim_to = c(10, 15, 20)) {
  dimsets <- base::expand.grid(list("from" = dim_from, "to" = dim_to))
  rownames(dimsets) <- apply(dimsets, 1, function(x) paste(x, collapse = "_"))
  return(dimsets)
}
create_umap_names <- function(reduction = "pca", dimsets = NULL) {
  umap_names <- sapply(rownames(dimsets), function(x) {
    paste0("umap_", reduction, "_", dimsets[x, "from"], "_", dimsets[x, "to"])
  })
  return(umap_names)
}
create_clustering_names <- function(reduction = "pca", dimsets = NULL, resolution = NULL) {
  clustering_names <- sapply(rownames(dimsets), function(x) {
    paste0("RNA_snn_", reduction, ".", dimsets[x, "from"], ".", dimsets[x, "to"], "_res.", resolution)
  })
  return(clustering_names)
}

run_umap_on_dimsets <- function(
  seurat_obj,
  reduction = "pca",
  dimsets = NULL
) {
  message(sprintf('Running %i UMAPs on %s reduction', nrow(dimsets), reduction))
  umap_names <- create_umap_names(reduction = reduction, dimsets = dimsets)

  for(dimsetId in rownames(dimsets)){
    message(sprintf("Running UMAP on %s dimensions %s", reduction, str_replace(dimsetId, "_", ":")))

    seurat_obj <- RunUMAP(
      object = seurat_obj,
      reduction = reduction,
      dims = dimsets[dimsetId, "from"]:dimsets[dimsetId, "to"],
      reduction.key = paste0(umap_names[dimsetId], "_"),
      reduction.name = umap_names[dimsetId]
    )
  }
  return(seurat_obj)
}

run_clustering_on_dimsets <- function(
  seurat_obj,
  reduction = "pca",
  dimsets = NULL,
  resolution = NULL
) {
  message(sprintf('Running %i clustering steps on %s reduction', nrow(dimsets), reduction))
  clustering_names <- create_clustering_names(reduction = reduction, dimsets = dimsets, resolution = resolution)

  for(dimsetId in rownames(dimsets)){
    message(sprintf("Running clustering on %s dimensions %s", reduction, str_replace(dimsetId, "_", ":")))
    seurat_obj <- FindNeighbors(object = seurat_obj, reduction = reduction, dims = dimsets[dimsetId, "from"]:dimsets[dimsetId, "to"])
    seurat_obj <- FindClusters(object = seurat_obj, resolution = resolution, cluster.name = clustering_names[dimsetId])
  }
  return(seurat_obj)
}

plot_comp_x_to_y <- function(obj, reduction.name='pca', compxtoy=1:6, colBy='nCount_RNA', colorscale = viridis::scale_colour_viridis()){
  #plot components 1:12 against each other (from pca, svd, mds...)
  #eg: plot_comp1to12(seurat_atac, reduction.name = 'pca_all_5000_most_var', colBy = 'sample_name')
  gg_df <-obj@reductions[[reduction.name]]@cell.embeddings[,compxtoy] %>% 
    as.data.frame %>% rownames_to_column('id')
  names(gg_df) <- c('id', paste0('PC_', compxtoy))
  gg_df <- as.data.frame(cbind(gg_df, obj@meta.data))
  gg_list <- list()
  j=1
  for(i in seq(min(compxtoy), max(compxtoy), by=2)){
    if(is.null(colBy)){
      gg_list[[j]] <- ggplot(gg_df) + geom_pointdensity(aes_string(x=sprintf('PC_%i',i), y=sprintf('PC_%i',i+1)), size = 0.5) + colorscale + theme_minimal()
    }else{
          gg_list[[j]] <- ggplot(gg_df) + geom_point(aes_string(x=sprintf('PC_%i',i), y=sprintf('PC_%i',i+1), colour=colBy), size = 0.5) + colorscale + theme_minimal()
    }
    j=j+1
  }
  #prime_numbers = prime_factorization(length(gg_list))
  #nrows = sum(prime_numbers[1:ceiling(length(prime_numbers)/2)])
  #ncols = ceiling(length(gg_list)/nrows)#sum(prime_numbers[(ceiling(length(prime_numbers)/2)+1):length(prime_numbers)])           
  nrows= round(sqrt(length(gg_list)))
  ncols = ceiling(length(gg_list) / nrows)
  ggpubr::ggarrange(plotlist = gg_list, nrow=nrows, ncol=ncols, labels='', label.x = 1/4, common.legend = T) 
}

```

```{r loaddata, echo=F, eval=T}

SObj <- readRDS(file.path(analysis_dir, config$Seurat_filtered))
#if(!is.null(config$scParam$vars_to_regress)){
#  stopifnot(regress_var %in% colnames(SObj@meta.data))
#}
pt_size = get_point_size(ncol(SObj))
samples<-names(table(SObj@meta.data$sample))

batch_col <- "lib"
batch_col_label <- batch_col

if (integrate) {
  batch_info_path <- file.path(analysis_dir, config$integration$batch_info_file)
  message('Adding batch information to SObj meta.data')
  batch_df = suppressMessages(read_csv(batch_info_path, col_types = 'cc'))
  stopifnot(identical(colnames(batch_df), c('orig.ident', 'batch')))
  stopifnot(!anyNA(batch_df))
  stopifnot(all(batch_df$orig.ident %in% SObj$sample))
  stopifnot(all(SObj$sample %in% batch_df$orig.ident))
  meta_df = SObj@meta.data[, !colnames(SObj@meta.data) == 'batch']
  SObj@meta.data = meta_df %>%
    rownames_to_column('temp_bc') %>%
    dplyr::left_join(batch_df, by = c('sample' = 'orig.ident')) %>%
    as.data.frame %>%
    column_to_rownames('temp_bc')

  batch_col <- "batch"
  batch_col_label <- "Batch"
}

dimsets <- create_dimset_table(dim_from = config$scParam$PCfrom, dim_to = config$scParam$PCto)

```

## Normalization, variable feature detection and PCA

```{r pca, include=F, message=FALSE, eval=T}

SObj <- NormalizeData(SObj, normalization.method = "LogNormalize", scale.factor = 10000)

# Perform cell cycle scoring

if (config$species == "homo sapiens") {
  s.genes <- cc.genes$s.genes
  g2m.genes <- cc.genes$g2m.genes
} else if (config$species == "mus musculus") {
  s.genes <- tools::toTitleCase(tolower(cc.genes$s.genes))
  g2m.genes <- tools::toTitleCase(tolower(cc.genes$g2m.genes))
} else {
  stop("Supplied species not supported for cell cycle scoring")
}

cc_features <- list(
  "S" = s.genes,
  "G2M" = g2m.genes
)

SObj <- AddModuleScore(SObj, features = cc_features, ctrl = 100, name = 'signature_module_')
SObj <- AddModuleScore_UCell(SObj, features = cc_features, name = "_UCellModule")

if (integrate) {
  SObj[["RNA"]] <- split(SObj[["RNA"]], f = SObj$batch)
}

SObj <- FindVariableFeatures(SObj, selection.method = "vst", nfeatures = config$scParam$VarFeat)

# Scaling with cell cycle regression

if (config$scParam$regress_cc) {
  if (config$scParam$regress_cc_score == "ModuleScore") {
    cc_vars <- paste0("signature_module_", names(cc_features))
  } else if (config$scParam$regress_cc_score == "UCell") {
    cc_vars <- paste0(names(cc_features), "_UCellModule")
  } else {
    stop("Invalid parameter provided for 'regress_cc_score'")
  }

  SObj <- ScaleData(SObj, features = VariableFeatures(SObj), vars.to.regress = cc_vars)
} else {
  # Scaling without cell cycle regression
  SObj <- ScaleData(SObj, features = VariableFeatures(SObj))
}

SObj <- RunPCA(SObj, features = VariableFeatures(object = SObj), npcs = 50)

SObj = run_umap_on_dimsets(
  SObj,
  reduction = 'pca',
  dimsets = dimsets
)

SObj = run_clustering_on_dimsets(
  SObj,
  reduction = 'pca',
  dimsets = dimsets,
  resolution = cluster_resolution
)

```

```{r plot var features, warning=F, echo = F}

VariableFeaturePlot(SObj)

```

### Decide on which PCs to use

We need to determine the primary sources of heterogeneity in the dataset and decide which or how many
PCs we use for the subsequent analysis. 

#### Elbow plot

PCs before the "elbow" in the percentage explained variance explain most of the non-random variation in the dataset.

```{r elbow, echo=F}

ElbowPlot(SObj, ndims = 50)

```

#### PC scores scatterplots

Plotting the PC scores against each other can help to reveal, which components capture 'noise': these will have a mostly round shape around zero.

```{r PC scores scatterplots, echo=F, eval=T}

#around 30 components seem to contain nonrandom variation
plot_comp_x_to_y(SObj, compxtoy = 1:20, colBy = NULL)
plot_comp_x_to_y(SObj, compxtoy = 21:40, colBy = NULL)

```

#### PC sequencing depth correlation

Sometimes, clustering can be improved by omitting the first PC if it strongly correlates with sequencing depth.

```{r ncount correlation, echo = F}

SObj$nCount_RNA_q95 = jj::jj_cap_vals(SObj$nCount_RNA, cap_top = 'q95')
plot_comp_x_to_y(SObj, compxtoy = 1:4, colBy = 'nCount_RNA_q95')
SObj$nCount_RNA_q95 = NULL
#plot_comp_x_to_y(SObj, compxtoy = 41:50, colBy = 'nCount_RNA')

```

```{r multiple_integration, include=F, message=FALSE, eval=T}

if (integrate) {
  for (integration_method in integration_methods) {
    SObj <- IntegrateLayers(
      object = SObj, method = integration_method_wrappers[[integration_method]],
      orig.reduction = "pca", new.reduction = integration_method,
      verbose = FALSE
    )

    SObj = run_umap_on_dimsets(
      SObj,
      reduction = integration_method,
      dimsets = dimsets
    )

    SObj = run_clustering_on_dimsets(
      SObj,
      reduction = integration_method,
      dimsets = dimsets,
      resolution = cluster_resolution
    )
  }

  SObj <- JoinLayers(SObj)
}

```

## UMAP for the data set

Seurat offers several non-linear dimensional reduction techniques. We chose  here UMAP to visualize the dataset.
If integration was chosen, UMAPs of integrated data for the different methods will be shown in addition to UMAP on PC space.

`r if(integrate) {paste0("Integration with methods: ", paste(integration_methods, collapse = ", "))}`

```{r generate_umap_plots, message=FALSE, echo=F}

reductions <- c("pca", integration_methods)
umap_reduction_plots <- list()

for (reduction in reductions) {
  umap_reduction_plots[[reduction]] <- list()
  umap_names <- create_umap_names(reduction = reduction, dimsets = dimsets)
  clustering_names <- create_clustering_names(reduction = reduction, dimsets = dimsets, resolution = cluster_resolution)

  for (dimsetId in rownames(dimsets)) {
    umap_plot_batch <- jj_plot_features(
      SObj,
      reduction = umap_names[[dimsetId]],
      meta_features = batch_col,
      pt.size = pt_size,
      do_shuffle = TRUE,
      return_gg_object = TRUE
    )[[1]] + theme(legend.position = "right") +
      guides(color = guide_legend(title = batch_col_label))

    umap_plot_clusters <- jj_plot_features(
      SObj,
      reduction = umap_names[[dimsetId]],
      meta_features = clustering_names[[dimsetId]],
      pt.size = pt_size,
      do_shuffle = TRUE,
      return_gg_object = TRUE
    )[[1]] + theme(legend.position = "right") +
      guides(color = guide_legend(title = paste0("Clusters\n(res ", cluster_resolution, ")")))

    umap_plots_combined <- umap_plot_batch + umap_plot_clusters
    umap_reduction_plots[[reduction]][[dimsetId]] <- umap_plots_combined

    # plt_name <- paste0("umap_", reduction, "_dims_", dimsetId, ".png")
    # ggsave(file.path(analysis_dir, config$outputpath$fig, "tmp", plt_name), height = 6, width = 12)
  }
}

```

```{r plot_umaps, results='asis', fig.dim = c(12, 6), echo = FALSE, warning = F}

for (reduction in names(umap_reduction_plots)) {
  cat(paste0("### Reduction: ", reduction, " {.tabset}"))
  cat("\n\n")
  for (dimset in names(umap_reduction_plots[[reduction]])) {
    cat("#### ", dimset, "\n")
    print(umap_reduction_plots[[reduction]][[dimset]])
    cat("\n\n")
  }
}

```


```{r save data, include=F}
#save back to Seurat filtered to avoid using too much disk space
#config$Seurat_dimred = file.path(dirname(config$Seurat_rawall), 'SeuratObj_merged_dimred.rds')

saveRDS(SObj, file=file.path(analysis_dir, config$Seurat_filtered))
#write_yaml(config, file=params$config)

```

## SessionInfo

```{r sessionInfo}
sessionInfo()
```
