---
title: "GSEA of differential expression between conditions"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
params:
  config: config.yaml
  config_id: ""
  date: !r Sys.Date()
---

# Notes

**Column explanation:**

* gene_set_lib: Gene set library
* pathway: pathway or gene set from library
* pval: unadjusted p-value
* padj_fdr: FDR-adjusted p-value
* NES: Normalized enrichment score. GSEA pathway enrichment score normalized by gene set size.
A positive normalized enrichment score means upregulation in first group compared to the second group in
the comparison
* gene_set_coverage: Fraction of genes in the gene set that could be tested for differential expression
* size: Size of the gene set (after removing untested genes)

```{r date, echo=FALSE}

#params <- list(config = "config.yaml", config_id = "CD4-CD8_A", date = Sys.Date())

today <- params$date
sprintf('Report run on %s.', today)

```

```{r load_libraries, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

libload <- c("stringr", "dplyr", "yaml", "knitr")

for (i in libload) {
    library(i, character.only = T)
}

# Custom functions

source("functions.R")

```

```{r setup, echo=FALSE, warning=FALSE }

set.seed(7)

root_directories <- get_root_directories()
analysis_dir <- root_directories[["analysis"]]

config <- yaml.load_file(params$config)

```

```{r parameters, echo=FALSE}

config_id <- params$config_id
config_key <- paste0("GSEA_conditions", ifelse(nzchar(config_id), "_", ""), config_id)

```

```{r load_data,eval=T,echo=F}

fgseaResults <- readRDS(file.path(analysis_dir, config[[config_key]]$results))

```

# Regulated pathways/gene sets by identity and comparison

```{r show_fgsea_results, eval=T, echo=F, results='asis'}

print(today)
cat("\n\n")

for (ident in names(fgseaResults)) {
    cat(paste0("## ", ident))
    cat("\n\n")
    comparisons <- unique(fgseaResults[[ident]]$comparison)

    for (comp in comparisons) {
        cat(paste0("### ", comp))
        cat("\n\n")

        currTable <- fgseaResults[[ident]] %>%
            filter(comparison == comp) %>%
            select(!all_of(c("comparison", "ES", "padj", "log2err", "padj_fdr", "leadingEdge"))) %>%
            rename(padj_fdr = padj_fdr_by_lib) %>%
            filter(padj_fdr <= 0.1) %>%
            arrange(padj_fdr) %>%
            mutate(pval = format(pval, digits = 3, scientific = TRUE)) %>%
            mutate(padj_fdr = format(padj_fdr, digits = 3, scientific = TRUE))

        tableUp <- currTable %>% filter(NES >= 0)
        tableDown <- currTable %>% filter(NES < 0)

        cat("#### Upregulated")
        cat("\n\n")

        if (nrow(tableUp)) {
            print(kable(tableUp))
            cat("\n\n")
        }

        cat("#### Downregulated")
        cat("\n\n")

        if (nrow(tableDown)) {
            print(kable(tableDown))
            cat("\n\n")
        }
    }
}

```

# SessionInfo

```{r sessionInfo}

sessionInfo()

```

