---
title: "Marker genes"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
params:
  config: config.yaml
  config_id: ""
  date: !r Sys.Date()
---


```{r date, echo=FALSE}

#params <- list(config = "config.yaml", date = Sys.Date())

today <- params$date
sprintf('Report run on %s.', today)

```

```{r load libraries, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

libload<-c("rprojroot", "yaml","tidyverse", "Seurat", "future","jj","knitr") 
invisible(sapply(libload, library, character.only = T))

# Custom functions

source("functions.R")

```

```{r setup, echo=FALSE, warning=FALSE}

set.seed(7)

root_directories <- get_root_directories()
analysis_dir <- root_directories[["analysis"]]

config <- yaml.load_file(params$config)

# parallelization in Seurat with future
# see https://satijalab.org/seurat/articles/future_vignette.html
plan("multisession", workers = 6)
options(future.globals.maxSize= 2000*1024^2) #otherwise error that future max size is exceeded

# bibliography
try(knitr::write_bib(c(.packages()), file.path(analysis_dir, config$outputpath$report, "R.bib")))
stopifnot(file.exists(file.path(analysis_dir, config$markers)))

```

```{r parameters, echo=FALSE}

config_id <- params$config_id
config_key <- paste0("markers", ifelse(nzchar(config_id), "_", ""), config_id)

```

```{r default config,echo=FALSE}
#Seurat/ArchR file from the config
obj_use = 'Seurat_filtered'
umap_use = config$scParam$dimred #paste(c('umap', config$scParam$PCdims), collapse = '_')
clusters = setIfNull(config$scParam$clusters, 'RNA_snn_res.0.2')
marker_file = setIfNull(config[[config_key]], '')
make_plots = TRUE #this will be set to false if no markers are available - the workflow runs through without error but generates no plots then
```

## Marker genes overview

```{r load marker file, echo = F, warning=T, message=T}
message('Reading marker file from path ', file.path(analysis_dir, marker_file))
markers = read_markers(file.path(analysis_dir, marker_file))
if(length(markers[[1]]) > 0){
  mdf = head(jj_list_to_df(markers, long=F, fill_with = ''), 10)
names(mdf) = sprintf("%s (%i)", names(mdf), sapply(markers, length))
kable(mdf, caption = 'First 10 entries for each marker category with (n marker genes)')
}

if(length(markers) == 1){
  markers = unlist(markers)
  names(markers) = NULL
}else{
  markers = setNames(unlist(markers, use.names=F),rep(names(markers), lengths(markers)))
}
if(length(markers) == 0){
  make_plots = FALSE
  markers_avail = 0 #make vector non-empty since sizes of code chunks depend on its length
  gmat_use = jj_initialize_df(2,2, row.names=c('a','b')) #initialize dummy df since sizes of code chunks depend on its rownames
}
```

```{r loaddata,eval=make_plots,echo=F}
obj <- readRDS(file.path(analysis_dir, config[[obj_use]]))
pt_size = get_point_size(ncol(obj))
dr_df = jj_get_reduction_coords(obj, umap_use)
stopifnot(clusters %in% colnames(dr_df))
```


```{r filter marker genes, echo=F, eval = make_plots}
features_avail = rownames(obj)

markers = switch(which_case(features_avail),
                 "lowercase"=tolower(markers),
                 "titlecase"=tools::toTitleCase(tolower(markers)),
                 "uppercase"=toupper(markers))

markers_avail = markers[markers %in% features_avail]
if(!all(markers %in% features_avail)){
  message(sprintf('%i/%i marker genes present in the object. Missing genes:\n%s', length(markers_avail), length(markers), paste(markers[!markers %in% features_avail], collapse = '\n')))
}
if(length(markers_avail) == 0){
  warning('No marker features were found. Skipping plots.')
  make_plots = FALSE
  markers_avail = 0 #make vector non-empty since sizes of code chunks depend on its length
  gmat_use = jj_initialize_df(2,2, row.names=c('a','b')) #initialize dummy df since sizes of code chunks depend on its rownames
}

marker_groups <- unique(names(markers))

```

## Gene expression per group

Use the groups defined in the config: `r clusters`

```{r cluster umap, echo = F, message = F, warning = F, eval = make_plots}
plot_df = jj_get_reduction_coords(obj, redname = umap_use)
jj_plot_features(reduction = plot_df, meta_features = clusters, cont_or_disc = 'd', label_col='white', label_type = 'geom_label', pt.size = pt_size, do_shuffle = T)
```

### Heatmap

If clusters are defined, get the mean expression per cluster and cluster on the scaled mean values in a heatmap. 

```{r plot heatmap, echo = F, message = F, warning = F, eval = make_plots, fig.width=12, fig.height=5+length(markers_avail)/10}
gmat = LayerData(obj, assay = "RNA", layer = "data")
gmat = gmat[unique(markers_avail), , drop=F]
gmat_use = gmat[apply(gmat, 1, function(x) n_distinct(x) > 1),,drop=F ]
try(print(jj_plot_heatmap(gmat_use, features_use = rownames(gmat_use), group_vec = obj@meta.data[[clusters]], transpose = T)))
```

### Dotplot

The dotplot additionally shows the fraction of cells within each group having at least one detected fragment for the markers.

```{r plot dotplot, echo = F, message = F, warning = F, eval = make_plots, fig.width=12, fig.height=5+length(rownames(gmat_use))/10}
try(print(plot_dotplot(gmat_use, features_use = rownames(gmat_use), group_vec = obj@meta.data[[clusters]], cluster_columns = T, cluster_rows = T, scale_data = T, cap_top = 'q98')))
```

## UMAP coloured by gene expression

```{r make ggplots, message = F, echo = F, eval = make_plots}
titles = NULL
if(!is.null(names(markers_avail)[1])){
  titles = paste0('Marker: ', names(markers_avail))
}
gg = jj_plot_features(seurat_obj = obj, reduction = umap_use, features = markers_avail, cap_top = 'auto', pt.size = pt_size, do_shuffle = T, return_gg_object = T)
for(i in seq_along(gg)){
  gg[[i]] = gg[[i]] +
    viridis::scale_color_viridis(labels = scales::label_number(accuracy = 0.01)) +
    labs(title= titles[i], colour=DefaultAssay(obj))
}

```

```{r plot markers, results='asis', echo = FALSE, eval = make_plots}

for (group in marker_groups) {
  curr_markers <- markers_avail[names(markers_avail) == group]
  curr_plots <- gg[names(markers_avail) == group]

  cat(paste0("### Marker group: ", group, " {.tabset}"))
  cat("\n\n")

  for (i in 1:length(curr_plots)){
    cat("#### ", curr_markers[i],"\n")
    print(curr_plots[[i]])
    cat('\n\n')
  }
}

```

## UMAP coloured by gene expression, ordered

```{r make ggplots ordered, message = F, echo = F, eval = make_plots}
titles = NULL
if(!is.null(names(markers_avail)[1])){
  titles = paste0('Marker: ', names(markers_avail))
}
gg = jj_plot_features(seurat_obj = obj, reduction = umap_use, features = markers_avail, cap_top = 'auto', pt.size = pt_size, do_order = T, return_gg_object = T)
for(i in seq_along(gg)){
  gg[[i]] = gg[[i]] +
    viridis::scale_color_viridis(labels = scales::label_number(accuracy = 0.01)) +
    labs(title= titles[i], colour=DefaultAssay(obj))
}

```

```{r plot markers ordered, results='asis', echo = FALSE, eval = make_plots}

for (group in marker_groups) {
  curr_markers <- markers_avail[names(markers_avail) == group]
  curr_plots <- gg[names(markers_avail) == group]

  cat(paste0("### Marker group: ", group, " {.tabset}"))
  cat("\n\n")

  for (i in 1:length(curr_plots)){
    cat("#### ", curr_markers[i],"\n")
    print(curr_plots[[i]])
    cat('\n\n')
  }
}

```

## Violin plots per group

Show the distribution of gene expression values for each group. The circles below the violins indicate the fraction of cells with >0 transcripts in the group (black area). The mean gene expression is shown as red line.

```{r make violins, message = F, echo = F, eval = make_plots}
gg = list()
titles = NULL
if(!is.null(names(markers_avail)[1])){
  titles = paste0('Marker: ', names(markers_avail))
}

#replace dash with underscore since ggplot throws error when - is in feature name
rownames(gmat) = gsub('-', '_', rownames(gmat))
markers_avail = gsub('-', '_', markers_avail)
for(i in seq_along(markers_avail)){
    gg[[i]] = jj_plot_sparse_by_group(gmat, 
                                      gene_plot = markers_avail[i], 
                                      group_vec =  obj@meta.data[[clusters]], 
                                      plot_zero_fraction = T, plot_mean = T, order = F, cap_top = 'q99',
                                      custom_colors = jj_get_jj_colours(gtools::mixedsort(unique(obj@meta.data[[clusters]]))))
}

```

```{r plot violins, results='asis', echo = FALSE, eval = make_plots}

for (group in marker_groups) {
  curr_markers <- markers_avail[names(markers_avail) == group]
  curr_plots <- gg[names(markers_avail) == group]

  cat(paste0("### Marker group: ", group, " {.tabset}"))
  cat("\n\n")

  for (i in 1:length(curr_plots)){
    cat("#### ", curr_markers[i],"\n")
    print(curr_plots[[i]])
    cat('\n\n')
  }
}

```

## SessionInfo

```{r sessionInfo}
sessionInfo()
```

