---
title: "Preprocessing"
author: "Inma Hernandez & Nicholas Strieder, modified by Malte Simon"

output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
params:
  config: config.yaml
  date: !r Sys.Date()
---

```{r notes, eval = F, include = F}

#The function Read10x is not flexible with the input file names
#Expected are matrix.mtx, features.tsv and barcodes.tsv (.gz also ok) within the filtered_feature_bc_matrix folders
#It might be necessary to rename files especially when using downloaded data with custom names (eg. from GEO). Replace substrings in all files e.g. with
#for file in *; do mv "${file}" "${file/GSM6668927_P2_T./}"; done

```


```{r date, echo=FALSE}

#params <- list(config = "config.yaml", date = Sys.Date())

today <- params$date
sprintf('Report run on %s.', today)

```

```{r load libraries,echo=FALSE, warning=FALSE, message = F}

knitr::opts_chunk$set(echo = TRUE)

# libraries
libload<-c("rprojroot", "yaml", "Seurat","scales","tidyverse", "future","knitr", "DropletUtils", "DropletQC")
invisible(sapply(libload, library, character.only = T))

source("functions.R")

```


```{r setup, echo=FALSE, warning=FALSE}

set.seed(7)

root_directories <- get_root_directories()
analysis_dir <- root_directories[["analysis"]]
data_dir <- root_directories[["data"]]

config <- yaml.load_file(params$config)

# parallelization in Seurat with future
# see https://satijalab.org/seurat/articles/future_vignette.html
plan("multisession", workers = 6)
options(future.globals.maxSize= 2000*1024^2) #otherwise error that future max size is exceeded

#create directories if not present
sapply(config$outputpath, function(x) dir.create(file.path(analysis_dir, x), recursive = TRUE))

# bibliography
try(knitr::write_bib(c(.packages()), file.path(analysis_dir, config$outputpath$report, "R.bib")))

```

```{r default config, echo = F}

config$filters$min_cells = setIfNull(config$filters$min_cells, 5)

```

```{r custom functions, eval=T,echo=F,message=F,warning=F}

loadndemux<-function(datdir){
    # hashtags: anti human:A0251-A0260,A0262-A0265,A0276,B0251-B0260,C0251-C0260,C0262-C265
    #           anti mouse:A0301-A0315,B0301-B310,C0301-C0310
    hashtags <- c(paste("A0",c(251:260,262:265,276,301:315),sep=""),
                  paste("B0",c(251:260,301:310),sep=""),
                  paste("C0",c(251:260,262:265,301:310),sep=""))
    # LoadData
    # if named vector is supplied, the cellids are prefixed with the ID
    if(length(names(datdir))==0){stop("No named vector supplied")}
    matrices <- Read10X(data.dir = datdir)
    # Demux
    #To demultiplex, we will follow the ["Demultiplexing with hashtag oligos (HTOs)" pipeline from 
    #Seurat](https://satijalab.org/seurat/v3.0/hashing_vignette.html):

    SObj<-CreateSeuratObject(counts = matrices$`Gene Expression`, min.cells= config$filters$min_cells ,min.features=1)
    # Add HTO as an independent assay
    feattab<-suppressMessages(read_tsv(file.path(datdir,"features.tsv.gz"),col_names=F))
    colnames(feattab)<-c("ID","name","type")

    feattab <- feattab %>% filter(grepl("Antibody Capture",type))

    # deal with modified matrices names
    ABnames<-rownames(matrices$`Antibody Capture`)
    if(!all(ABnames%in%pull(feattab,name))){
        if(all(str_replace(ABnames,"[.].","")==pull(feattab,name))){
            feattab$name<-ABnames
          }else{
            names(ABnames)<-str_replace(ABnames,"[.].","")
            ABnames<-ABnames[pull(feattab,name)]
            if(all(str_replace(ABnames,"[.].","")==pull(feattab,name))){
                feattab$name<-ABnames
            }else{
              message("AB names are not matching")
            }
          }
    }

    
    # select all IDs which match to hashtags from Biolegend, see above to update
    hashtab <- feattab %>% filter(ID %in% hashtags)
    # select all IDs which do not match to hashtags and use as CSFs
    feattabf <- feattab %>% filter(!ID %in% hashtags)
    
    if(length(pull(hashtab,name="ID"))>0){
        if(max(matrices$`Antibody Capture`[hashtab$name,])>0){
            # there are hash antibodies in this library, 
            # but maybe not all hashAB were used in this library
        
    # need to remove rows with rather low counts otherwise 'HTODemux' will fail
    # for HTO there should be at least 0.1 % of the median reads from all other hashes
    hashSums<-Matrix::rowSums(matrices$`Antibody Capture`[hashtab$name,])
    hashthres<-median(hashSums)*0.001
    # keep must be vector of names!!!
    keephash<-hashtab$name[which(hashSums>hashthres)]
    message(paste("processing library:" ,names(datdir),sep=""))
    message(paste("Will keep hashes with reads above threshold: ",
                  paste(names(keephash),collapse=", "),sep=""))
    if(length(which(hashSums<hashthres))>0){
    message(paste("Will discard hashes with reads below threshold: ",
                  paste(names(which(hashSums<hashthres)),collapse=", "),sep=""))}

    SObj[["HTO"]] <- CreateAssayObject(counts = matrices$`Antibody Capture`[keephash,])

    #sanity check
    if(!all(rownames(SObj[["HTO"]]) == str_replace_all(hashtab$name, "_", "-"))) {
        stop("Error in selection of Hash AB names")
    }
    SObj  <- NormalizeData(SObj, assay = "HTO", normalization.method = "CLR")
    # Demultiplex
    
    SObj <- HTODemux(SObj, assay = "HTO", positive.quantile = 0.99)
    }else{print("No reads for Hashing AB found")}
    }else{print("no hashing info available")}

    if(length(pull(feattabf,name="ID"))>0){
      if(max(matrices$`Antibody Capture`[feattabf$name,])>0){
    # Add ADT as an independent assay
    SObj[["ADT"]] <- CreateAssayObject(counts = matrices$`Antibody Capture`[feattabf$name,])
    SObj <- NormalizeData(SObj, assay = "ADT", normalization.method = "CLR",margin=2)
      }else{print("No reads for Feature AB found")}
    }else{print("No feature info available")
    }
    return(SObj)
}


loadd<-function(datdir){
    # LoadData
    # if named vector is supplied, the cellids are prefixed with the ID
    if(length(names(datdir))==0){stop("No named vector supplied")}
    matrices <- Read10X(data.dir = datdir)
    # Demux
    #To demultiplex, we will follow the ["Demultiplexing with hashtag oligos (HTOs)" pipeline from 
    #Seurat](https://satijalab.org/seurat/v3.0/hashing_vignette.html):

    SObj<-CreateSeuratObject(counts = matrices, min.cells= config$filters$min_cells, min.features = 1)
    return(SObj)
}


ridgeplot<-function(SObju){
    
kable(table(SObju$HTO_classification.global), caption = 'HTO classification global')
kable(table(SObju$HTO_classification), caption = 'HTO classification fine')
Idents(SObju) <- "HTO_maxID"
pd<-RidgePlot(
            object = SObju, 
              assay = "HTO", 
              features = rownames(SObju[["HTO"]]), 
                ncol = 3
              )
print(pd)
}


dimred_fun<-function(SObj, type = 'pca'){
    SObj_subset <- subset(
                       SObj,
                         subset = HTO_classification.global == "Negative",
                         invert = TRUE
                         )
    DefaultAssay(SObj_subset) <- "HTO"
    SObj_subset <- SObj_subset %>%
      ScaleData(features = rownames(SObj_subset), verbose = TRUE) %>%
        RunPCA(features = rownames(SObj_subset[["HTO"]]), approx = FALSE)
    if(type == 'tsne'){
      SObj_subset = RunTSNE(SObj_subset, dims = 1:2, perplexity = 100,check_duplicates = FALSE)
    }else if(type == 'umap'){
      SObj_subset = RunUMAP(SObj_subset, dims = 1:2) #does not make much sense with just two dimensions
    }
    pd <- DimPlot(SObj_subset, reduction = type, group.by = "HTO_classification.global")
     print(pd)
}

#fails if old cellranger output is used: gene_barcode_matrix/ instead of feature_barcode_matrix/ -> tryCatch
run_droplet_qc = function(path_to_data, out_name){
  outs_dirs = dirname(path_to_data)
  nf_list = list()
  for(i in seq_along(outs_dirs)){
      barcodepattern="gene_barcode_matrix$|feature_barcode_matrix$|filtered_feature_bc_matrix$"
      barcodespath <- dir(outs_dirs[i],pattern=barcodepattern)
      nf_temp = tryCatch({nuclear_fraction_tags(
      outs = outs_dirs[i],
      tiles = 100, cores = 6)}, error = function(x){
        nuclear_fraction_tags(
          bam = file.path(outs_dirs[i], 'possorted_genome_bam.bam'),
          bam_index = file.path(outs_dirs[i], 'possorted_genome_bam.bam.bai'),
          barcodes = file.path(barcodespath, 'barcodes.tsv.gz'),
          tiles = 100, cores = 6)})
    nf_temp$barcode = paste(names(path_to_data)[i], rownames(nf_temp), sep = '_')
    nf_list[[names(path_to_data)[i]]] <- nf_temp
  }
  nf_df = dplyr::bind_rows(nf_list)
  write_csv(nf_df, file = out_name)
}

```

## Introduction

HTO are cross-linked with antibodies that bind to ubiquitous cell surface proteins. We then multiplexed all batches and processed them into the same 10X Chromium Chip Channel ([GEM well](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/glossary)), hence sequencing together HTO and single-cell gene expression reads (cDNA). Finally, we can detect inter-batch doublets as those cell barcodes that have high counts of two or more HTO.

The objective of this notebook is to demultiplex the barcodes (cells) back to its original batch to discard doublets and negative cells.

## Loading and demutlipexing the data 

```{r find cellranger data,eval=T,echo=F,message=T,warning=F}

# # Samples from inputdir
# samples<-list.dirs(file.path(data_dir, config$inputpath),full.names=F,recursive=F)
# sampnames<-sub("_run_[0-9]*","",sub("p[0-9A-z]{3}_","",samples))
# if(any(grepl("_sc[A-z]{3,4}_",samples))){
#    sampnames<-sub("sc[RTC][NCS][ARP].{0,1}_","",sampnames)
# }
# sampnames<-gsub("_","-",sampnames)
# names(samples)<-sampnames
# 
# 
# # cellranger outputpath
# if(file.exists(file.path(data_dir, config$inputpath,samples[1],"outs", 'filtered_feature_bc_matrix'))){
#   path_to_data <- sapply(samples, 
#                          function(x) file.path(data_dir, config$inputpath,
#                                     x,"outs","filtered_feature_bc_matrix"))
# }else{
#   path_to_data <- sapply(samples, 
#                          function(x) file.path(data_dir, config$inputpath,
#                                     x,"outs","filtered_gene_bc_matrices"))
# }

message('Searching for input files and generating sample names with maxlength=20')
path_to_data = find_input_files(
  file.path(data_dir, config$inputpath),
  dir_pattern = 'filtered_feature_bc_matrices|filtered_feature_bc_matrix|filtered_gene_bc_matrices|filtered_gene_bc_matrix',
  samples_to_exclude = config$libs_exclude,
  ocm_output = config$on_chip_multiplexing
)

message('Cellranger output paths: ')
message(paste(paste(names(path_to_data), path_to_data, sep = ": "), collapse='\n'))
# harmony_template_df = data.frame(orig.ident = names(path_to_data), batch = names(path_to_data))
# harmony_template_path <- file.path(config$outputpath$table, "harmony_template.csv")
# message('Saving batch template for Harmony (which would treat each sample as one batch) in file:\n', harmony_template_path)
# write_csv(harmony_template_df, file = file.path(analysis_dir, harmony_template_path))

```

```{r Print cellranger paths, message = FALSE, echo = F}

#message again for command line (message printed to console when message = FALSE)
message('Cellranger output paths: ')
message(paste(paste(names(path_to_data), path_to_data, sep = ": "), collapse='\n'))

## Harmony 
if(!is.null(config$harmony_csv)){
message('Found template for Harmony in file:\n', config$harmony_csv)}
#There is a template for harmony.csv in the tables folder, which would treat each sample as one batch

```

```{r load data,eval=T,echo=F,message=T,warning=F}

# load data into seurat, used named vector to have samples with names

SObjl<-list()
if(config$cellhash){
  message('Cellhashing is set to `TRUE`')
    # cellhasing used
   for ( samp in names(path_to_data)){
        SObjl[[samp]]<-loadndemux(path_to_data[samp])
   }        

}else{
    # no cellhasing
   for (samp in names(path_to_data)){
       SObjl[[samp]]<-loadd(path_to_data[samp])
   }
}

# save data as rds
Seurat_rawobj<-list()
for(samp in names(SObjl)){
    outname <- file.path(config$outputpath$data,
                         paste0("SeuratObj_raw_",samp,".rds"))
    # datapath=file.path(config$outputpath,"data")
    # if(!dir.exists(file.path(analysis_dir, datapath))){
    #     dir.create(file.path(analysis_dir, datapath))
    #    print("created data path")
    # }
    message(paste0('Saving ', outname))
    saveRDS(SObjl[[samp]],file = file.path(analysis_dir, outname))
    Seurat_rawobj[[samp]]<-outname
}

config$Seurat_rawobj <- Seurat_rawobj

# join data as rds
if (length(SObjl) > 1) {
  SObjf <- merge(x = SObjl[[1]], y = SObjl[2:length(SObjl)])

  # Merging Seurat object automatically creates layers which we do not want at this point, so join them for now

  SObjf[["RNA"]] <- JoinLayers(SObjf[["RNA"]])
} else {
  SObjf = SObjl[[1]]
}

#sanity check on the species specification in config.yaml
if(which_case(rownames(LayerData(SObjf, assay = "RNA", layer = "counts"))) == 'titlecase' & config$species == 'homo sapiens'){
  warning('Gene names have title case, but species was set to `homo sapiens`. Wrong species specified?')
}else if(which_case(rownames(LayerData(SObjf, assay = "RNA", layer = "counts"))) == 'uppercase' & config$species == 'mus musculus'){
  warning('Gene names are uppercase, but species was set to `mus musculus`. Wrong species specified?')
}

```

## DropletQC

```{r DropletQC, eval = T, echo = F}

# TODO: This does not work right now, as it searches for the bam files inside the cellranger count matrix folders

dropletqc_file = file.path(analysis_dir, config$outputpath$table, "dropletQC_nuclear_fraction.csv")
if(!file.exists(dropletqc_file)){
  message('Running DropletQC')
  try(run_droplet_qc(path_to_data = path_to_data, out_name = dropletqc_file))
}

if(file.exists(dropletqc_file)){
  nf_df = read_csv(dropletqc_file)
  meta_df = SObjf@meta.data
  meta_df$barcode = rownames(meta_df)
  meta_df = dplyr::left_join(meta_df, nf_df, by = 'barcode')
  #then get the nCount nuclear_fraction
  empty_drops = meta_df[, c('nuclear_fraction','nCount_RNA', 'barcode')]
  empty_drops = empty_drops[complete.cases(empty_drops), ] 
  #add the empty droplet classification
  empty_drops = tryCatch({identify_empty_drops(empty_drops)},
                         error = function(x){
                           warning('An error occured in `identiy_empty_drops`: ', x)
                           empty_drops})
        
  #the damaged cell classification can only be added after celltype identification
  #empty_drops$celltype = 'Treg' #use singler results here in SingleR.Rmd
  #empty_drops = identify_damaged_cells(empty_drops[, c('nuclear_fraction', 'nCount_RNA', 'cell_status', 'celltype')])
  SObjf$nuclear_fraction = empty_drops$nuclear_fraction[match(meta_df$barcode, empty_drops$barcode)]
  if('cell_status' %in% colnames(empty_drops)){
    SObjf$cell_status = empty_drops$cell_status[match(meta_df$barcode, empty_drops$barcode)]
  }
}

```

```{r emptyDrops, eval = F, echo = F}

# TODO: estimate empty droplets using DropletUtils, it appears that this currently does not work:
# Error in .compute_ambient_stats(mat, totals, lower = lower) : no counts available to estimate the ambient profile (in p123 A test)

##this requires to read in the raw_feature_matrix instead of the filtered one without setting any filters!
library(DropletUtils)
#https://github.com/MarioniLab/DropletUtils/issues/14
#https://bioconductor.org/packages/devel/bioc/vignettes/DropletUtils/inst/doc/DropletUtils.html
#raw feature barcode matrix should be used
ass_df = LayerData(SObjf, assay = "RNA", layer = "counts")
set.seed(100) #seed needed for reproducibility
#lower is 100 by default -> barcodes with lower count are assumed to be empty droplets
e.out <- emptyDrops(ass_df, lower = 100)  
e_use = e.out[!is.na(e.out$FDR), ]
cell_barcodes <- rownames(e_use)[e_use$FDR <= 0.01]
writeLines(cell_barcodes, 'emptyDrops_predictedCells.txt')

# We recommend making some diagnostic plots such as the total count against the negative log-probability. 
# Droplets detected as cells should show up with large negative log-probabilities _or_ very large total counts (based on the knee point reported by `barcodeRanks`).
plot(e.out$Total, -e.out$LogProb, col=ifelse(is.cell, "red", "black"),
    xlab="Total UMI count", ylab="-Log Probability")
```

## Merged dataset overview

```{r print object overview, echo = F}

SObjf

```

```{r print n cells, echo = F}

table(SObjf$orig.ident)

```

```{r save data,eval=T,echo=F}

outname <- file.path(config$outputpath$data, "SeuratObj_raw_merged.rds")
#paste("SeuratObj_raw_merged_",today,".rds",sep="")
message(paste0('Saving ', outname))
saveRDS(SObjf,file = file.path(analysis_dir, outname))
config$Seurat_rawall <- outname
write_yaml(config,file=params$config)

```

`r if(config$cellhash) {"## Cell hashing visualization"}`
`r if(config$cellhash) {"### Ridge plots {.tabset}"}`
    
```{r ridgeplot,fig.height=10, fig.width=14,echo=FALSE,message=F,warning=F, eval = config$cellhash, results='asis'}

for (samp in names(SObjl)) {
  if("HTO" %in% names(SObjl[[samp]])) {
        #print(paste("Ridge plot for sample: ",samp,sep=""))
        cat("#### ", samp, "\n")
        ridgeplot(SObjl[[samp]])
        cat('\n\n')
  }
}

```

`r if(config$cellhash) {"### Heatmaps {.tabset}"}` 
`r if(config$cellhash) {"In this heatmap, each column represents a 10X barcode, and each row the expression of a batch-specific HTO. We can visualize singlets, doublets and negative cells (ie not labelled by any HTO). We strive for a similar number of cells across batches, as this increases our ability to identify doublets."}`

```{r Heatmap,eval= config$cellhash,,echo=FALSE,message=F,warning=F, results='asis'}

for (samp in names(SObjl)) {
  if("HTO" %in% names(SObjl[[samp]])) {
    #print(paste("Heatmap for sample: ",samp,sep=""))
    cat("#### ", samp, "\n")
    pd<-HTOHeatmap(SObjl[[samp]], assay = "HTO", ncells = 5000)
    print(pd)
    cat('\n\n')
  }
}

```

`r if(config$cellhash) {"### tSNE {.tabset}"}` 

`r if(config$cellhash) {"We can visualize the tSNE embedding of cells in HTO space. A clear separation of clusters would indicate a good labeling:"}`  
  
```{r tSNE, eval = config$cellhash,echo=FALSE,message=F,warning=F,  results='asis'}

for (samp in names(SObjl)) {
  if("HTO" %in% names(SObjl[[samp]])){
    #print(paste("tSNE for sample: ",samp,sep=""))
    cat("#### ", samp, "\n")
    dimred_fun(SObjl[[samp]], type = 'pca')
    cat('\n\n')
  }
}

```

## SessionInfo()

```{r sessionInfo,echo=T}
sessionInfo()
```
