---
title: "SingleR"
author: "Malte Simon, custom function by Niklas Beumer"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
params:
  config: config.yaml
  date: !r Sys.Date()
---

```{r date, echo=FALSE}

today <- params$date
sprintf('Report run on %s.', today)

```

```{r load libraries,include=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# libraries
libload<-c("rprojroot", "yaml","Seurat", "tidyverse", "future", "jj", "SingleR")
invisible(sapply(libload, library, character.only = T))

# Custom functions

source("functions.R")

```

```{r setup, echo = F, message = F, warning= F, error=T}

set.seed(7)

root_directories <- get_root_directories()
analysis_dir <- root_directories[["analysis"]]

config <- yaml.load_file(params$config)

# parallelization in Seurat with future
# see https://satijalab.org/seurat/articles/future_vignette.html
plan("multicore", workers = 6)
options(future.globals.maxSize= 2000*1024^2) #otherwise error that future max size is exceeded

try(knitr::write_bib(c(.packages()), file.path(analysis_dir, config$outputpath$report, "R.bib")))

```

```{r default config,echo=FALSE}
reference_dir = config$singlerRefpath
ref_list_custom = NULL
obj_use = 'Seurat_filtered'
umap_use = config$scParam$dimred #paste(c('umap', config$scParam$PCdims), collapse = '_')
fast_mode = FALSE #speed up singler predictions by summarising reference samples by their label
```

```{r custom functions, echo = F, message = F, warning= F, error=T}
run_singler <- function(seur_obj, ref_dataset){
  # This function applies SingleR to a Seurat object. Cells are annotated to cetagories
  #   in the label.main slot of the reference data set.
  # seur_obj: The Seurat object to use.
  # ref_dataset: A SingleR reference data set.
  # Dependencies: Seurat, testit, SingleR (The latter must be installed but need 
  #   not be loaded)
  # Value: The updated Seurat object.
  # Author: Niklas Beumer
  
  # Convert the Seurat object to a SingleCellExperiment.
  seur_obj_for_singler <- as.SingleCellExperiment(seur_obj, assay = "RNA")
  #stopifnot(identical(colnames(seur_obj), colnames(seur_obj_for_singler)))
  
  # Run SingleR
  singler_res <- SingleR::SingleR(test = seur_obj_for_singler,
                                  ref = ref_dataset,
                                  labels = ref_dataset$label.main)
  return(singler_res)
}

get_singler_label = function(singler_res){
  # Re-name cells that are unclassified.
  singler_res$pruned.labels <- replace(singler_res$pruned.labels, which(is.na(singler_res$pruned.labels)), "Unclassified")
  
  # Append SingleR's annotations to the Seurat object.
  #stopifnot(identical(colnames(seur_obj), rownames(singler_res)))
  cell_type_order <- sort(unique(singler_res$pruned.labels))
  pred_vec <- factor(singler_res$pruned.labels, levels = cell_type_order)
  
  # Return the updated Seurat object.
  return(pred_vec)
}

```

```{r loaddata,eval=T,echo=F}
SObj <- readRDS(file.path(analysis_dir, config[[obj_use]] ))
pt_size = get_point_size(ncol(SObj))
stopifnot(umap_use %in% Reductions(SObj))
sample_names <-names(table(SObj@meta.data$sample))
```


```{r get references, include = F}

# Get the reference data sets.
if (config$species == 'homo sapiens') {
  # ref_list = list(
  #   novershtern <- celldex::NovershternHematopoieticData(cell.ont = "all"),
  #   blueprint <- celldex::BlueprintEncodeData(cell.ont = "all"),
  #   hpca <- celldex::HumanPrimaryCellAtlasData(cell.ont = "all"),
  #   dice <- celldex::DatabaseImmuneCellExpressionData(cell.ont = "all"),
  #   monaco <- celldex::MonacoImmuneData(cell.ont = "all")
  # )
  ref_list = list(
    novershtern = 'NovershternHematopoieticData.RDS',
    blueprint = 'BlueprintEncodeData.RDS',
    hpca = 'HumanPrimaryCellAtlasData.RDS',
    dice = 'DatabaseImmuneCellExpressionData.RDS',
    monaco = 'MonacoImmuneData.RDS'
  )
} else if (config$species == 'mus musculus') {
  # ref_list = list(
  #     immgen <- celldex::ImmGenData(cell.ont = "all"),
  #     mouse_rna_seq <- celldex::MouseRNAseqData(cell.ont = "all")
  # )
  ref_list = list(
      immgen = 'ImmGenData.RDS',
      mouse_rna_seq = 'MouseRNAseqData.RDS'
  )
} else {
  stop('config$species should be one of homo sapiens or mus musculus')
}

ref_list <- lapply(ref_list, function(x) file.path(reference_dir, x))

if (!is.null(ref_list_custom)) {
  ref_list <- c(ref_list, ref_list_custom)
}

message('Using the following reference datasets: ', paste(names(ref_list), collapse = ', '))
ref_list <- lapply(ref_list, read_rds)

if(fast_mode){
  for(i in seq_along(ref_list)){
    mean_mat = jj_summarize_sparse_mat(assays(ref_list[[i]])[[1]], ref_list[[i]]$label.main)
    ref_list[[i]] = SummarizedExperiment(assays = list(logcounts = mean_mat), colData = data.frame(label.main=colnames(mean_mat), label.fine=colnames(mean_mat)))
  }
}
```

```{r run singler, include = F, message = F}
gg = gg2 = gg3 = singler_res_list = list()
counter = 1 
for(i in names(ref_list)){
  message(sprintf('%i/%i: %s', counter, length(ref_list), i))
  # Set a seed since the processes in this code chunk involves random sampling.
  set.seed(7)
  
  # Perform SingleR annotation.
  singler_res_list[[i]] = run_singler(seur_obj = SObj, ref_dataset = ref_list[[i]])
  SObj[[i]] <- get_singler_label(singler_res_list[[i]])
  
  # Generate a colour scale for the annotated cell types.
  cell_type_order <- sort(unique(SObj@meta.data[, i]))
  n_celltypes <- length(cell_type_order)
  ann_col_palette <- c(sample(rainbow(n_celltypes - 1), size = n_celltypes - 1), "black")
  names(ann_col_palette) <- cell_type_order
  gg[[i]] = jj_plot_features(SObj, reduction = umap_use, meta_features = i, pt.size = pt_size, do_shuffle = T, return_gg_object = T)[[1]]
  gg2[[i]] = jj_plot_features(SObj, reduction = umap_use, meta_features = i, facet_by = i, pt.size = pt_size, do_shuffle = F, use_pointdensity = T, return_gg_object = T)[[1]]
  gg3[[i]] = jj_plot_categorical_by_group(SObj@meta.data, feature_column = i, group_column = 'sample', add_text = T)
  counter = counter+1
}
```


```{r save seurat, echo = F}
saveRDS(SObj, file.path(analysis_dir, config[[obj_use]] ))
```

## SingleR predictions on UMAP {.tabset}

For this automatic cell type annotation, we use SingleR. SingleR assesses the correlation between a cell's gene expression and reference gene expression profiles for particular cell types. There are different collections of reference profiles. Currently, we use references provided along with SingleR in the celldex package.

```{r plot umap, results='asis', echo = FALSE, warning=F, fig.height=5, fig.width=8 + 2 * ceiling(length(cell_type_order) / 20) - 2}
#fig.height=2 * ceiling(n_celltypes / 4), fig.width=12 + 4 * ceiling(length(cell_type_order) / 20) - 4}
for (i in names(ref_list)){
  cat("### ", i,"\n")
  #cat('UMAP coloured by SingleR prediction')
  print(gg[[i]] + ggtitle('UMAP coloured by SingleR prediction'))
  #cat('SingleR label density')
  print(gg2[[i]] + ggtitle('SingleR label density'))
  #cat('SingleR label frequency summarized by group')
  print(gg3[[i]] + ggtitle('SingleR label frequency summarized by group'))
  #cat('SingleR prediction scores')
  print(SingleR::plotScoreHeatmap(singler_res_list[[i]], main = 'SingleR prediction scores'))
  #cat('SingleR delta distribution')
  print(SingleR::plotDeltaDistribution(singler_res_list[[i]]) + ggtitle('SingleR delta distribution'))
  cat('\n\n')
}
```

## SessionInfo()

```{r sessionInfo,echo=T}
sessionInfo()
```
